---
title: "Behavioral Data Science Week 4 - Factor Analysis"
author: "Jonathan Gragg"
date: "9/19/2021"
output: pdf_document
---

## Summary

The initial goal was to determine whether or not 17 variables in the employee survey could be averaged into one variable. I first glimpsed the data and turned all of the factors into a numerical data value in order to evaluate. My initial choice was to run the factors through a scree plot and see what that tells me. It led to the conclustion that there needed to be a least 3 factors. That would contradict the initial hypothesis that these variables could become 1 factor so my initial conclusion was that these variables could not be averaged into one variable.

After that I fit a model to both an Orthogonal analysis and a Oblique analysis. In an orthogonal you are testing whether or not the factors should remain independent while in an oblique you are testing whether or not the factors should become 1 factor. In both findings I was able to see that there was some variables that would point to factor 2 and factor 1. This confirmed my conclusion that these variables could not be averaged into 1 score.

## Analysis

loading in relevant Libraries

```{r}
library(tidyverse)
library(ggplot2)
library(stringr)
library(psych)
```

loading in the data

```{r}
data <- read.csv("survey_results_public.csv")

#selecting the columns that we will be working with
df <- data %>% select(ProblemSolving:ChangeWorld) %>% drop_na()

#converting all the of the column values into a numerical rating scale
df[df=='Strongly disagree'] <- 1
df[df=='Disagree'] <- 2
df[df=='Somewhat agree'] <- 3
df[df=='Agree'] <- 4
df[df=='Strongly agree'] <- 5

columns = 1:17

for (i in columns){
  df[,i] = as.integer(df[,i])
}

glimpse(df)
```

```{r}
#Is a FA analysis appropriate?

#if the p value is small then FA analysis is appropriate
cortest.bartlett(df)

#if overall MSA is over a certain threshold then yes
KMO(df)
```

```{r}
scree(df)
```

```{r,warning=FALSE}
#Number of factors?
fa.parallel(df)
```

Based on the parallel analysis scree plot I would say that there should be 3 factors and not 1. That would lead me to believe that these metrics cannot be averaged into one.

```{r}
#orthogonal rotate
fa2 <- fa(df, nfactors = 3, rotate = "varimax", covar = FALSE)
print(fa2$loadings, cutoff = 0.001, digits = 3)
```

```{r}
fa2$communalities
```

```{r}
fa.diagram(fa2)
```

```{r}
#oblique rotate
fa3 <- fa(df, nfactors = 3, rotate = "promax")
print(fa3$loadings, cutoff = 0.001, digits = 3)
```

because there are variables that are higher in MR2 than in MR1 we can conclude that there needs to be more than 1 factor.

```{r}
#look at correlation between latent factors
fa3$Phi
```

There is very little correlation between the factors which would lead me to believe that my original conclusion was correct that these variables should not be averaged into 1.

```{r}
fa.diagram(fa3)
```

```{r}
##PC vs. FA
pc1 <- principal(df, nfactors = 3, rotate = "none", covar = FALSE)
class(pc1)
fa.diagram(pc1)
```

```{r}
alpha(df)
```

