---
title: "Behavioral Data Science Week 5 - IRT and GRM"
author: "Jonathan Gragg"
date: "9/26/2021"
output: pdf_document
---

## Summary

Based on the IRT model I would say that the distribution looks very similar for 1 IRT model across all questions. The GRM model shows a little disparity however and looks to be 2-3 different distribution shapes. That would lead me to confirm what I discovered last week where I suggested there needed to be 3 factors for these variables.

## Analysis

loading in relevant Libraries

```{r}
library(tidyverse)
library(ggplot2)
library(stringr)
library(psych)
library(GGally)
library(mirt)
library(ltm)
library(lavaan)
```

loading in the data

```{r}
data <- read.csv("survey_results_public.csv")

#selecting the columns that we will be working with
df <- data %>% dplyr::select(ProblemSolving:ChangeWorld) %>% drop_na()

#converting all the of the column values into a numerical rating scale
df[df=='Strongly disagree'] <- 1
df[df=='Disagree'] <- 2
df[df=='Somewhat agree'] <- 3
df[df=='Agree'] <- 4
df[df=='Strongly agree'] <- 5

columns = 1:17

for (i in columns){
  df[,i] = as.integer(df[,i])
}

glimpse(df)
```

```{r}
#Try 1-factor
fa1 <- fa(df, nfactors = 1, rotate = "none")
print(fa1$loadings, cutoff = 0.001, digits = 3)
```

```{r}
fit1PL <- mirt(df, model = 1, itemtype = "Rasch")
plot(fit1PL, type = 'trace', facet_items = FALSE)
```

```{r}
confirm_mod1 <- "Endurance =~ ProblemSolving + BuildingThings + LearningNewTech + BoringDetails + JobSecurity + DiversityImportant +
AnnoyingUI + FriendsDevelopers + RightWrongWay + UnderstandComputers + SeriousWork + InvestTimeTools + WorkPayCare + 
KinshipDevelopers + ChallengeMyself + CompetePeers + ChangeWorld"
fit1 <- lavaan::cfa(confirm_mod1, data = df,std.lv = TRUE)
summary(fit1, fit.measures = T,standardized = T)
```

```{r}
facScores = fscores(fit1PL)

head(facScores)
```

```{r}
summary(facScores)
```

```{r}
faScores = psych::fa(df, 1, rotate = "promax")$scores

summary(faScores)
```

```{r}
cor(faScores, facScores)
```

```{r}
grmMod = mirt(df, model = 1, itemtype = "graded")

coef(grmMod, simplify = TRUE, IRTpars = TRUE)
```

```{r}
plot(grmMod, type = 'trace', 
     facet_items = TRUE, main = "GRM")
```

