---
title: "Text Evaluation - Warmup"
author: "Jonathan Gragg"
date: "11/17/2021"
output: pdf_document
---

# Summary

In this analysis we will be taking words people have used to describe their managers to then help determine whether they are a bad manager or a good manager. We'll give each manager a sentiment score to see which ones have predominantly negative sentiment scores vs predominantly positive sentiment scores.

# Sentiment Analysis

```{r}
# Loading in all required packages
library(tidyverse)
library(tm)
library(tidytext)
library(wordcloud2)
library(sentimentr) 
library(lexicon)
library(magrittr)
```

```{r}
# loading in data
load("week13.RData")
glimpse(week13)
```

## Tidying

```{r}
# limiting data and taking out the special characters to be able to run sentiment analysis
week13Tidy = week13 %>% 
  select(id, manager, describeManager) %>% 
  ungroup() %>%
  mutate(describeManager = unlist(.$describeManager), 
         describeManager = gsub("^c\\(|\\)$", "", .$describeManager)) 
```

## Sentiment score model 1

```{r}
# basic sentinent analysis to determine if there are more positve or more negative words
employeeSentiment = week13Tidy %>%
  unnest_tokens(tbl = ., output = word, input = describeManager) %>% 
  group_by(id) %>% 
  inner_join(get_sentiments("bing")) %>% 
  count(sentiment) %>% 
  spread(sentiment, n, fill = 0) %>% 
  mutate(sentiment = positive - negative)

employeeSentiment
```

## Sentiment score model 2

```{r}
# getting more complex by taking the mean number of positive versus negative for each manager then taking the difference
managerSentiment = week13Tidy %>%
  unnest_tokens(tbl = ., output = word, input = describeManager) %>% 
  group_by(manager, id) %>% 
  inner_join(get_sentiments("bing")) %>% 
  count(sentiment) %>% 
  spread(sentiment, n, fill = 0) %>% 
  group_by(manager) %>% 
  summarize(meanNegative = mean(negative), 
            meanPositive = mean(positive), 
            sentiment = meanPositive - meanNegative)

managerSentiment
```

```{r}
# validating the worst score has negative words and max score has positive scores
worstManager = managerSentiment$manager[which.min(managerSentiment$sentiment)]

bestManager = managerSentiment$manager[which.max(managerSentiment$sentiment)]

week13$describeManager[week13$manager == worstManager][sample(1:5, 1)]
```

```{r}
week13$describeManager[week13$manager == bestManager][sample(1:5, 1)]
```

## Sentiment score model 3

```{r}
# using a lexicon set for a sentiment score then creating a score for each of our managers off of that score
rinkSent = sentiment(week13Tidy$describeManager, polarity_dt = lexicon::hash_sentiment_jockers_rinker)

week13Tidy %>% 
  mutate(element_id = 1:nrow(.)) %>% 
  left_join(., rinkSent, by = "element_id") %>% 
  group_by(manager) %>% 
  summarize(meanSent = mean(sentiment)) %>%
  arrange(-meanSent)
```

